---
title: "Penalised Gibbs Estimation: Example (v.1.0)"
author: "Tuomas Rajala"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 6-variate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction
This document provides demonstrates the use of the R-package `PenGE` for estimating significant interactions in a multivariate Gibbs model using Group Lasso automatic variable selection. 


Start by making sure required packages are installed:

```{r, results=FALSE}
library(PenGE)
library(spatstat)
```


## Data
Let's see if we can discover interactions in 6-variate `lansing` data set.

```{r, fig.width=7, fig.height=6}
data(lansing)
plot(split(lansing), cex=.5)
```

## Fitting the model
The data is $p=6$ variate point pattern. 

### Bare minimum

The bare minimum is to define the interaction ranges:

```{r}
p <- 6 # we have six species
rv <- c(0.025, 0.05) # use these for everything
r1 <- rep( list(rv) , p) # intra-type interaction ranges
r2 <- rep( list(rv), p * (p-1)/2) # inter-type
```

Then we need to compute the logistic regression pseudo-likelihood model matrix.

```{r, cache=TRUE}
Q <- make_Q_stepper_multi(x=lansing, ranges1 = r1, ranges2 = r2, border_r = 0.05)
```

Then we can estimate the parameters with group lasso (no Cross validation):

```{r, cache=TRUE}
fit <- fitGlbin_CV(Q)
```

We can look at the AIC path to get an idea if the model fits at all:

```{r}
with(fit$fullfit, plot(lambda , aic, xlim = rev(range(lambda) )))
```

Get the estimated interaction matrix, with signs. Default penalty level (when param `k` is not set) is ad-hoc rule-of-thumb: half-way between max. penalty and AIC minimum. 

```{r, fig.width = 6, fig.height=6}
M <- imatrix(fit, signed = TRUE)
image.imatrix(M, col = c("red","black","green"), cex.axis = 0.7)
```

Use CV to more objectively choose the penalty level.

### Cross validation
For cross validation we split the pattern into parts spatially. In order to have truly spatially independent datasets, we need to use eroded windows. This in turn leads to a loss of data. Let's accept 50% loss. How many quadrats can we have?

```{r}
r_border <- 0.05
win_side_len <- 1
nmax <- cv_border_solve_split(R_frac = r_border/win_side_len, acceptable_loss = 0.5)
nx <- round( sqrt(nmax) )
quads <- split_window(lansing$window, nx = nx, ny = nx)
```

Check split:
```{r, fig.width=6, fig.height=6}
plot(unmark(lansing), cex=.3)
for(q in quads) plot(erosion(w = q, r = r_border), add=T, fg=2)
```

Not that many. We can increase `nx` but at the cost of introducing correlation in the CV estimates. 


Next we need to fit the model again, this time several times for the different subsets. Current implementation uses `mclapply` from ``parallel` (adjust `mc.cores` to your resources):

```{r, cache=TRUE}
cvfit <- fitGlbin_CV(Q, quads = quads, mc.cores = 5)
```

Then we compute CV Risk as average of the quadrats' squared-residual sums. For Gibbs models, different residuals are available. Lots of numerical computations are hidden here, but by default something sensible should emerge. One can for example set at `dum_grid_dimyx = c(30,30)` :


```{r, cache=TRUE}
Rest <- residuals_cv(cvfit, x = lansing, mc.cores = 6)
```

The individual residual curves per CV-fit and the *inverse*-residual:

```{r, fig.width=7, fig.height=3}
inv <- Rest$by_resid$inverse
pen <- cvfit$lambda 

par(mfrow=c(1,2))

# The residual per penalty for each CV quadrat
plot(pen, inv[1,], "l", xlim = rev(range(pen)), ylim=range(inv))
for(i in 2:nx^2) lines(pen, inv[i,], col=i)
# Estimated Risk:
risk_inv <- Rest$ave$inverse
plot(pen, log(risk_inv), "l", xlim = rev(range(pen)))
```

The CV-risk optimal interaction matrix:

```{r, fig.width = 6, fig.height=6}
kopt <- which.min(risk_inv)
Mcv <- imatrix(cvfit, k = kopt, signed = TRUE)
image.imatrix(Mcv, cols = c(2,1,3), cex=.7)
```



## Further options

### Covariates



